---
title: "lab_12"
author: "Mitchell Hang"
date: "2023-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidytext)
library(tidyverse)
library(dplyr)
library(janitor)
library(lubridate)
library(rvest)
```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 10 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 10 words describe?

```{r}
#Load the data
md_attorney_sanctions <- read_csv("data/md_attorney_sanctions.csv")
```

```{r}
#Create list of unique words
unique_words <- md_attorney_sanctions |> 
  select(text) |>
  unnest_tokens(word, text)
```

```{r}
#Load stop words
data("stop_words")
```

```{r}
#Remove stop words from unique words list
unique_words |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**A1. The leading word is "failing" at 4.59%, and it makes sense that a word with a negative connotation would be leading in text related to misconduct. Other top words include "client," "attorney" and "conduct," which also makes sense given it is describing how these Maryland attorneys got into trouble and was likely related to how they interacted with clients.**

------------------------------------------------------------------------

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

```{r}
# Check the task 12 on Pre-lab 12 to see if you can use that code
trigrams <- md_attorney_sanctions |>
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  filter(!word3 %in% stop_words$word) |>
  mutate(trigram = paste(word1, word2, word3, sep=" ")) |>
  group_by(trigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**A2. The top trigram is "attorney trust account," which appeared 343 times. This is referring to a bank account that an attorney is supposed to use to store money that they are handling for their client. The fact that it appeared so often indicates there was probably a lot of fraud occurring related to this money.**

------------------------------------------------------------------------

**Q3.** Let's drop back down to more traditional text analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. What do you think the answer produced by your code suggests? What else could you do to try and clarify the most important reasons attorneys get sanctioned?

```{r}
md_attorney_sanctions |>
  filter(str_detect(text, "attorney trust account")) |>
  group_by(fiscal_year) |>
  tally(sort=TRUE)
```

**A3. The term "attorney trust account" appeared the most in the fiscal year of 2021 at 37 times. This suggests that people may have had funds in their attorney trust accounts in relation to the pandemic, which may have led to lawyers stealing these funds. I would like to do more text searches of phrases related to theft and other types of misconduct in order to learn more about their reasons.**
