---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
options(scipen=999)

library(tidyverse)
library(tidycensus)
library(lubridate)
```

## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

-   Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

```{r}
pg_calls <- read_csv("data/prince_georges_2022_overdoses.csv")

pg_calls <- pg_calls %>%
  mutate(
    date = date(datetime),
    week = week(datetime),
    month = month(datetime, label = TRUE)
  )
```

## Questions

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

-   A clear title that states the main idea/finding
-   Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
-   Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1. There do seem to be some significant differences between this bar chart and the one from the prelab, for many of the months seem to be flipped. While in the prelab August had the highest number of calls, in this chart August is third from the bottom. Similarly, in the prelab December was the second from the bottom, whereas it is the months with the highest amount of calls in the lab.**

```{r}
pct_by_month <- pg_calls 

pct_by_month <- pct_by_month %>%
  mutate(total = nrow(pct_by_month)) %>%
  group_by(month, total) %>%
  summarize(count = n())

pct_by_month <- pct_by_month %>%
  mutate(pct = (count/total)*100) %>%
  select(-total)

pct_by_month %>%
  ggplot() +
  geom_bar(aes(x=reorder(month, pct), weight=pct)) +
  coord_flip() +
  labs(
    title = "Percentage of Prince George's County 911 Overdose Calls by Month",
    x = "month", 
    y = "percent of total calls",
    caption = "Source: Prince George's County EMS"
  ) +
  theme_minimal()
```

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

-   A clear title that states the main idea/finding
-   Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
-   Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2. This graph includes a fair amount of spikes and drops, with no clear pattern being visible. The highest spike is in week 28 with 41 calls, and at the very end it dips quite a lot with only 4 calls in week 53. The percent of total calls mostly stays in the range of 1.2 to 2.0, and the spikes are when it gets as high as 2.2 to 2.9.**

```{r}
pct_by_week <- pg_calls %>%
  group_by(week) %>%
  summarize(calls = n())

pct_by_week <- pct_by_week %>%
  mutate(pct = (calls/1397)*100)

pct_by_week %>%
  ggplot() +
  geom_line(aes(x=week, y=pct)) +
  labs(
    title = "Percentage of Prince George's County 911 Overdose Calls by Week",
    x = "week", 
    y = "percent of total calls",
    caption = "Source: Prince George's County EMS"
  ) +
  theme_minimal()

```

**Q3.** A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis. For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid.

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable. A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables:

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1.  Generate a dataframe with the number of 911 calls for each zip code.
2.  Get data from the Census Bureau showing median household income for Maryland zip codes.
3.  Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4.  Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5.  Give it an appropriate title, source, and x and y axis titles.
6.  Add a label for each point that shows the zip code using geom_text() - see some examples of its use at <https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples>. Try to make the names as easy to read as possible by avoiding overlap.
7.  In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3. From looking at this graph, there foes appear to be some relationship between a zip code's median income and the number of overdoses calls. While this is not the case for every zip code, some of the zip codes with the highest median incomes, such as 20769 and 2062, have less than five reported calls. The zip codes with the most calls, 20745 and 20743, are both below \$70,000 for median income, putting them in the lowest quarter of the data as far as median income.**

```{r}
pg_calls_zip <- pg_calls %>%
  group_by(zipcode) %>%
  summarize(calls = n())

md_median_income_zip <- get_acs(geography = "zcta", variables = "B19013_001", state = "MD", year = 2019)

pg_join_income_calls <- pg_calls_zip %>%
  inner_join(md_median_income_zip, by = c("zipcode" = "GEOID"))

ggplot(pg_join_income_calls, aes(x=calls, y=estimate, label=zipcode)) +
  geom_point() +
  geom_text(hjust = 0, nudge_x = 0.05, check_overlap = TRUE) + 
  labs(
    title = "911 Overdose Calls in Prince George's County by Median Household Income",
    x = "number of overdose calls", 
    y = "median household income",
    caption = "Source: Prince George's County EMS, Census Bureau"
  ) +
  theme_minimal()
```
