---
title: "Reverse Engineering Project"
author: "Sasha Allen, Nick Elliot, Hannah Marszalek, Bode Ramsay"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, we are reverse engineering the story, [STORY NAME](STORY%20URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

```{r}
# Load required data
homicide_data <- read_csv("data/homicide-data.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 
homicide_data <- homicide_data %>%
  mutate(fixed_date = ymd(reported_date))

```

## Sentences to Engineer

In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

-   **Sentence text:** \"As Baltimore has seen a stunning surge of violence, with nearly a killing each day for the past three years in a city of 600,000, homicide arrests have plummeted.\"
-   **Analysis summary**: By filtering for only the killings that happened in 2015, 2016, or 2017 in Baltimore and then grouping by year, we were able to see that the total number per year was pretty close to 365, thus confirming their claim that there was nearly a killing each day. We also added a column showing the percent of the year for which there was a killing per day, thus further confirming this.

```{r}
# Put code to reverse engineer sentence here
past_three_years <- homicide_data %>%
  filter(city == "Baltimore") %>%
  mutate(year = year(fixed_date)) %>%
  filter(year == "2015" | year == "2016" | year == "2017") %>%
  group_by(year) %>%
  summarise(
    total_killings = n(),
    percent_of_year = total_killings/365 * 100
  )
  
# Display results of code below this codeblock
past_three_years
```

### Sentence 2

-   **Sentence text:** \"Of 50 of the nation\'s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis. In Chicago, the homicide arrest rate has dropped 21 percentage points, in Boston it has dropped 12 points and in St. Louis it is down 9.\"
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 3

-   **Sentence text:** \"Baltimore is also one of 30 cities that have seen an increase in homicides in recent years, with the greatest raw number increase in killings of any city other than Chicago, which has four times the population. While homicide rates remain near historical lows in most American cities, Baltimore and Chicago are now both seeing murder tallies that rival the early 2000s.\"
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

-   **Sentence text**: \"For most of the decade before 2015, Baltimore\'s annual homicide arrest rate hovered at about 40 percent. Since 2015, the arrest rate hasn\'t topped 30 percent in any year. And while most cities saw their arrest rates drop gradually, Baltimore\'s decline was sudden --- plummeting 15 percentage points in 2015, after Gray\'s death, the largest single-year drop for any city already solving less than half its homicides.\"
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

-   **Sentence text:** \"Homicides have soared in several neighborhoods since Gray\'s death. Sandtown-Winchester, where Gray died, has seen 22 more homicides in the three-year period since Gray\'s death than it did in the three years before he died. Southwest Baltimore saw its homicides rise by 35, and Greater Rosemont has seen 26 more since 2015.\"
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-
