---
title: "Data Analysis Project"
author: "Matthew Kiras, Hannah Marszalek, Bode Ramsay, Hayden Sweeney"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Introduction:** This data notebook was created to analyze the data in a repository on women's college basketball rosters, which can be found [here](https://github.com/Sports-Roster-Data/womens-college-basketball).

### Load libraries

```{r}
library(tidyverse)
library(janitor)
```

### Load data

This data had already been cleaned when we began working with it, so there was not much to do in this regard.

```{r}
roster_2020_21 <- read_csv("data/wbb_rosters_2020_21.csv")
roster_2021_22 <- read_csv("data/wbb_rosters_2021_22.csv")
roster_2022_23 <- read_csv("data/wbb_rosters_2022_23.csv")
roster_2023_24 <- read_csv("data/wbb_rosters_2023_24.csv")

#combined dataframe
all_rosters <- bind_rows(
  roster_2020_21,
  roster_2021_22,
  roster_2022_23,
  roster_2023_24
)

old_years_cleaned <- read_csv("data/years_cleaned.csv")
years_cleaned <- old_years_cleaned |>
  group_by(`year-clean`) |> 
  summarize(sum(count, na.rm = TRUE))
```

**Question 1:** What states produce the most women's basketball players?

**Summary:** As far as the total number of players over the course of the four seasons, Pennsylvania, Texas and New York have the highest numbers. We also looked at the percent change in the numbers of players between 2020 and 2023, which was interesting because Pennsylvania had a change that was less than 1%, showing how they have stayed consistently near the top of the list in terms of how many players they produce. One issue that exists with the total is that it does not account for the fact that there may be duplicate players between some of the years, but finding a way to fix this would be quite challenging. We did try to do this using the all_rosters dataframe, but as we had some issues with cleaning the season column, we found it was easier to calculate the years separately and then join them.

```{r}
#create player counts by state for each year
state_counts_2020 <- roster_2020_21 |>
  group_by(state_clean) |>
  summarize(count_2020 = n()) |>
  na.omit()

state_counts_2021 <- roster_2021_22 |>
  group_by(state_clean) |>
  summarize(count_2021 = n()) |>
  na.omit()

state_counts_2022 <- roster_2022_23 |>
  group_by(state_clean) |>
  summarize(count_2022 = n()) |>
  na.omit() |>
  filter(state_clean != "USVI")

state_counts_2023 <- roster_2023_24 |>
  group_by(state_clean) |>
  summarize(count_2023 = n()) |>
  na.omit()

#combine all years to see total counts
#calculate percent change between 2020 and 2023 for each state
all_years_state_counts <- state_counts_2020 |>
  left_join(state_counts_2021, by = "state_clean") |>
  left_join(state_counts_2022, by = "state_clean") |>
  left_join(state_counts_2023, by = "state_clean") |>
  mutate(total_players = count_2020 + count_2021 + count_2022 + count_2023) |>
  mutate(pct_change = ((count_2023 - count_2020)/count_2020)*100) |>
  arrange(desc(total_players))

all_years_state_counts
```

**Question 2:** Which power 5 conferences have the tallest teams for the current season?

**Summary:** For this question, we looked at the most recent roster data and filtered for only players from the power 5 conferences: SEC, ACC, Big Ten, Big 12, and Pac-12. We then grouped by team and found the average height for each team, which showed us that Texas has the highest average height for the 2023-24 season at 73.33 inches. UCLA and Indiana are the second and third tallest teams, and Clemson is the shortest at 69.93.

```{r}
#filter for only the power 5 conferences
#create a column to calculate average height per team
tallest_teams_2023 <- roster_2023_24 |>
  filter(conference == "SEC" | conference == "Big Ten" | conference == "ACC" | conference == "Big 12" | conference == "Pac-12") |>
  group_by(team) |>
  summarise(
    avg_height = mean(total_inches)) |>
  arrange(desc(avg_height))

tallest_teams_2023
```

**Question 3:** What is the breakdown of freshmen, sophomores, juniors, seniors, and grad students per each team? What percentage of each team do redshirts make up?

**Summary:** To answer this question, we took the roster from the current season and grouped the different years to see the breakdown by team. We also used the redshirt column to identify the percentage of redshirt players from each team. Two teams have a lineup comprised of more than 60% redshirts, Grand Valley St. and North Carolina Central, both of which have 9 redshirts. Biola has the third highest percentage at 58.8%, but it has the highest raw number of redshirts at 10. Of the list of 932 teams, seven teams are 50% or more composed of redshirts. While this is not a particularly high number of teams, it could be interesting to explore why some teams are more than half redshirts.

```{r}
#group the dataframe by team and get a count of each year
years_2023 <- roster_2023_24 |>
  select(team, year_clean) |>
  group_by(team, year_clean) |>
  count()

years_2023 <- years_2023 |>
  pivot_wider(names_from = year_clean, values_from = n)

years_2023[is.na(years_2023)] <- 0
  
#add a column for redshirts
redshirts <- roster_2023_24 |>
  select(team, redshirt) |>
  filter(redshirt==1) |>
  group_by(team) |>
  summarize(redshirts = n())

years_2023 <- years_2023 |>
  left_join(redshirts, by="team") |>
  clean_names()

#mutate a total column as well as percent columns
years_2023 <- years_2023 |>
  mutate(total = freshman+graduate_student+junior+senior+sophomore+fifth_year+na+sixth_year) |>
  mutate(pct_redshirts = (redshirts/total)*100) |>
  arrange(desc(pct_redshirts))

years_2023
```

**Question 4:** What school took in the most transfers? What school had the most players transfer out?

**Summary:** Old Dominion and Glenville St. took in the most transfers at 14 each, with three people transferring out of Old Dominion and zero transferring out of Glenville St. Oregon, Arizona, and Blinn College had the most people transfer out, with nine people transferring out of Oregon and eight people each transferring out of Arizona and Blinn College. Interestingly, none of these teams had coaching changes over the offseason, which makes us question what the reason was for so many players leaving. 806 schools received zero transfers, and 174 schools had zero people transfer out.

```{r}
#filter for only the players with something in the previous_school column
all_transfers_2023 <- roster_2023_24 |>
  filter(!is.na(previous_school_clean))

#create two different dataframes: one grouped by team and one grouped by previous_school
current_team_transfers_grouped <- all_transfers_2023 |>
  group_by(team) |>
  summarize(transferred_in = n())

previous_school_transfers_grouped <- all_transfers_2023 |>
  group_by(previous_school_clean) |>
  summarize(transferred_out = n()) |>
  rename("team" = previous_school_clean)

#combined the two dataframes
combined_transfers_2023 <- full_join(current_team_transfers_grouped, previous_school_transfers_grouped, by = "team")

combined_transfers_2023[is.na(combined_transfers_2023)] <- 0

combined_transfers_2023
```

**Question 5:** Which teams have the most in-state players?

**Summary:** Looking again at only the rosters from the power-5 conferences, the schools with the most in-state players are Texas, Michigan State, and Minnesota, with nine players at Texas and eight players at both Michigan State and Minnesota. The state of Texas produces a lot of D1 women's basketball players, as we learned from one of our previous questions, and it appears that a lot of that talent stays home. These results are quite interesting considering schools such as NC State, Rutgers, and Oregon State, all of which have a reputation for having a majority of in-state students, have less than three in-state players on their roster for this year. Of the 59 total teams, only 13 have five or more in-state players. This suggests that the teams from these conferences focus largely on out-of-state recruitment.

```{r}
#create a dataframe with only the rows where the team state and home state match
in_state_players <- roster_2023_24 |>
  filter(team_state == state_clean) 

#filter for only the power 5 conferences and group by team to get a count of in state players
in_state_players <- in_state_players |>
  filter(conference == "SEC" | conference == "Big Ten" | conference == "ACC" | conference == "Big 12" | conference == "Pac-12") |>
  group_by(team) |>
  summarize(count=n()) |>
  arrange(desc(count))

in_state_players
```
