---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

**A1: It looks like it's a website url that was saved as a csv as opposed to a previously-created csv.**

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2: There are 9776 records in this dataset. Each record provides information on a different earthquake, with information such as where/when it happened and what type of earthquake it was.**

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes <- earthquakes %>%
  arrange(desc(depth))
```

**A3: To see the deepest first, use the arrange function and specify that you want to arrange in descending order according to the depth column. The depth that shows up for the deepest one is 669.9820, and its magnitude is 4.2.**

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
mag_greater_than_6 <- earthquakes %>%
  filter(mag > 6)
```

**A4: You can use the filter function to keep only the desired results. In this case, there are 12 earthquakes that fit the desired criteria.**

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
mag_and_depth <- earthquakes %>%
  filter(mag > 6 & depth < 20)
```

**A5: You can use a filter with an "and" operator. There are five records that fit both criteria.**

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
mag_or_depth <- earthquakes %>%
  filter(mag > 6 | depth < 20)
```

**A6: You can use a filter with an "or" operator. There are 7413 records that fit either criteria.**

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
alaska <- earthquakes %>%
  filter(str_detect(place, "Alaska"))
```

**A7: You can use a string detect function on the "place" column to find all the earthquakes that mention the location as being in Alaska. 3423 earthquakes took place there.**

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
tremor_types <- earthquakes %>%
  group_by(type) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```

**A8: You can group_by and then count to find all possible types with the number of each. The first two most common tremors are earthquakes at 9540 and quarry blast at 129.**

------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
tremor_depth <- earthquakes %>%
  group_by(type) %>%
  summarize(avg_depth = mean(depth))
```

**A9: The average depth of the earthquake type is 25.28. It is interesting that some have negative values.**

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes <- earthquakes %>%
  mutate(hour = hour(time))
```

**A10: Using a function from the lubridate package, you can extract just the hour from the time column, similar to when we did this with the month in an earlier lab.**

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}
earthquakes <- earthquakes %>%
  mutate(serious = case_when(
    mag > 3 ~ "serious",
    .default = "not serious"
  ))
```

**A11: You can use the mutate function to create a new column and then use a case_when to label all records with a magnitude greater than 3 with the word "serious".**

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
time_earthquakes <- earthquakes %>%
  filter(serious == "serious" & type == "earthquake") %>%
  group_by(hour) %>%
  summarize(count_serious = n()) %>%
  arrange(desc(count_serious))
```

**A12:** **By filtering for only serious earthquakes and then grouping by hour, you can see a count of how many serious earthquakes occur each hour. The hour with the fewest serious earthquakes is 12 pm, and the one with the most is 2 am.**

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?

```{r}

```

```{r}

```

```{r}

```
